---
title: "Literate Analysis"
author: "Taylor"
format: html
editor: visual
---

## Introduction

### About this data
This dataset includes bottle chemistry data from the North Pole Environmental Observatory. The data can be found at [this link](https://arcticdata.io/catalog/view/doi:10.18739/A25T3FZ8X). The data was downloaded on 6/26/2023.

### Setup
```{r}
library(readr)
```

### Read data
```{r}
#| warning: false
#| mesage: false
bg_chem <- read_csv("data/BGchem2008data.csv")
```
Exploring data

```{r}
#| eval: false
#| echo: false
colnames(bg_chem)

str(bg_chem)

head(bg_chem)

summary (bg_chem)

unique(bg_chem$Date)
```

## Analysis



### Calculate summary statistics
```{r}
nitrate <- mean(bg_chem$NO3)
nitrite <- mean(bg_chem$NO2)
amm <- mean(bg_chem$NH4)
phos <- mean(bg_chem$P)

```


### Calculate mean Redfield ratio
```{r}
ratio <- (nitrate + nitrite + amm)/phos
```


### Plot Refield ratio
```{r}
plot(bg_chem$P, bg_chem$NO3 + bg_chem$NO2 + bg_chem$NH4)
```


## Conclusion

The Redfield ratio for this data is approximately `r round(ratio)`